Rheology of Cubic Blue Phases

### 2012-09-28 papers/bprheo/bprheo.tex
Updated bprheo draft
git-svn-id: https://ccpforge.cse.rl.ac.uk/svn/ludwig/trunk@1786 40876441-3496-0410-a613-a2fca35b2ae2

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.

As it might be easier to enter a discussion with the full picture in mind, we provide first a 
quantitative analysis of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picutre shows the date over time, whereas the graph at the bottom gives 
the same data versus accummulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

@@

Again, we present key aspects in a kind of overview before we 
address some features in more detail. Fig. \ref{bp1-rheo} shows the 
apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
as in Fig. \ref{bp2-rheo}.
For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
the disclination network occurs similar to and at the same critical shear rate as 
the one observed in BPII-2 but with an order structure that is a quasi 
two-dimensional.

@@

The Fourier analysis of the time-dependent deviatoric stress $\Pi_{xy}$ 
offers another perspective on this intricate flow behaviour at high and intermediate shear rates. 
The spectrum $X_\omega$ according to Eq. \ref{spectrum} is shown in Fig. \ref{bp1-spectrum}.
We observe a strong contribution of the first harmonic, which is directly related to the 
recurrence period $T$, shear rate $\dot{\gamma}$ and unit cell size $l_{u}$ 
via $\omega_0=2\pi/T=2\pi/(l_{u}\dot{\gamma})$.
As the shear rates we applied always differ by a factor 2 the Fourier coefficients of the 
individual runs with different shear rate coincide in such a way that the
the fundamental mode is the same frequency as the first overtone of the next lower shear rate
or the third overtone of the next-to-next lower shear rate 
(viz. peak positions at $\gd=7.81, 3.91$ and $1.95\e{-5}$).
Moreover, the magnitude of higher Fourier modes decreases monotonously.   
However, the inset shows that this characteristic frequency relation is violated 
for shear rates above $\gd=7.81\e{-5}$.
Although the velocity of the primary flow component in x-direcition, listed in Tab. \ref{tab1},
 is very well in accordance with the imposed shear rate, the fundamental mode appears 
now at the same frequency.
The contribution of higher Fourier modes increases and even exceeds that of
the fundamental mode, which gives the shear stress generally a more jittering appearance.
At the highest imposed shear rate where the breakup occurs the Fourier spectrum 
becomes simple again and features a dominating fundamental mode at the 
expected frequency just as described above for $\gd\le7.81\e{-5}$.
The breakup occurs at exactly the same flow rate as those of BPII, 
suggesting that this can be traced back to the rotational dynamics 
of the order parameter.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI at different shear rates: The main picture shows the frequency shift of the Fourier components in agreement with the imposed shear rate. The inset depicts the same analysis at higher shear rate, where the frequency shift is no longer observed. Instead the spectral contribution of higher modes grows until finally the breakup takes place. The frequency is given in units of $10^2\cdot 2\pi/T$ with $T$ as period of one cycle in LB time steps.}
\label{bp1-spectrum}
\end{figure}

@@

The rheological response of BPI at low shear rate which exhibits a
dynamic yield stress is strikingly different from that of BPII.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the backgound viscosity has been subracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.
The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces excert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

@@

@@

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Snapshots of BPI disclination network at low shear rate ($\dot{\gamma}=4.88\e{-6}$): Depicted is the transition from flow-induced, intertwined helices that undergo a recurring affine transformation (top left and centre at time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) to a distorted state (bottom centre at time step $t=1.7\e{6}$) and finally to an amorphous network (bottom right at $t=2.2\e{6}$). All pictures have the vie along the vorticity direction apart from the one at the top centre which shows the network in gradient direction.}
\label{bp1-low}
\end{figure*}

However, this mode of flow proves unstable as after a few cycles 
distortions appear which lead to further destabilisation.
Finally, the network loses any residual order and 
transforms into an amorphous state with virtually constant
yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LU.

@@

Intermediate shear rates in the sense of this study of BPI are characterised 
by the missing negative branch in the deviatoric stress. It is unclear 
why this feature is missing at higher flow rates. A possible explanation is that 
above a certain flow rate the network simply skips the comparably small 
window of negative stress in Fig. \ref{bp1-fe-yield} and finds a quicker 
route to get back on the positive branch.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
\caption{}
\label{bp1-1_bp1-2}
\end{figure}

@@

The complexity of flow-induced states and the seemingly chaotic but 
periodic movement of the BPI network raises the question how robust 
these results are with respect to the system size.
Although the pitch length of 64 LU and the equivalent of 32 LU per
unit cell provide enough resolution to track the motion of the 
network in flow it is not clear whether 64 unit cells in total are enough to 
mimick true bulk behaviour.

@@

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\caption{Apparent viscosity $\eta_{app}$ for 8 (solid lines) and 64 unit cells (dashed lines)
at $\gd=9.76\e{-6}$ and $1.95\e{-5}$. The shape of each curve differs largely, but the recurrence frequencies are hardly affected by the system size.}
\label{bp1-2uc4uc}
\end{figure}

>>>>

Simple shear flow was imposed by means of a Lees-Edwards boundary condition ~\cite{Wagner:2002} with
the top (bottom) half of the planes moving in positive (negative) x-direction and the 
velocity gradient along the y-direction.
We applied a correction for spurious momentum currents that emerge due to required interpolations across 
the Lees-Edwards planes \cite{Henrich:2012a}.
The shear rates were varied over more than two orders of magnitude from about 
$\gd=2.44\times \e{-6}$ to $6.25\times\e{-4}$ in LB units.
For clarity we classify various subregimes, two in case of BPII and three in case of BPI. 
They are BPII-1 ($\gd < 6.25\e{-4}$), BPII-2 ($\gd\ge 6.25\e{-4}$) and
BPI-1 ($\gd < 3.91\e{-5}$), BPI-2 ($3.91\e{-5}\le\gd< 6.25\e{-4}$) and BPI-3 ($\gd\ge 6.25\e{-4}$). 
Disclination lines are plotted as isosurface of the scalar order parameter $q$. Typical values 
are $q_s=0.18$ for BPI and $q_s=0.15$ for BPII.

@@

We start our discussion with BPII as its flow behaviour turns out to be somewhat simpler 
than that of BPI. BPII has simple cubic symmetry and the $\pi$-disclination lines intersect and form a characteristic network of nodes.
Ahead of the more detailed discussion we provide first a general overview of the flow behaviour at all applied shear rates.
Fig. \ref{bp2-rheo} shows the apparent viscosity, which is defined as the 
deviatoric stress normalised to the isotropic background viscosity: 
$\eta_{app}=\Pi_{xy}/\eta\,\gd$.
A numerical value of $\eta_{app}=1$ corresponds to total shear thinning without 
any additional contribution from the liquid crystal.
The top picutre shows the date over time, whereas the graph at the bottom gives 
the same data versus accummulated strain.
For all but the highest flow rates $\eta_{app}$ oscillates sinusoidally, which
is entirely because of the periodic breakup and reconnecting of the network in shear flow. We refer to this regime as BPII-1.

@@

We present again key aspects in a kind of overview before we 
address some features in more detail. Fig. \ref{bp1-rheo} shows the 
apparent viscosity $\eta_{app}$ over time and strain for the same shear rates
as in Fig. \ref{bp2-rheo}.
For shear rates in regime BPI-3, i.e. $\gd\ge6.25\e{-4}$ a breakup of 
the disclination network occurs similar to and at the same critical shear rate as 
the one observed in BPII-2 but with an order structure that is a quasi 
two-dimensional.

@@

In order to gain a better understanding of the complex oscillatory behaviour in regime BPI-2 
we performed a Fourier analysis of the deviatoric stress $\Pi_{xy}$ according to Eq. \ref{spectrum}. 
The spectrum is shown in Fig. \ref{bp1-spectrum}.

There are two aspects which are worth mentioning. First of all, we observe a strong contribution of 
the fundamental mode for the lowest two shear rates and a relatively small contribution of higher harmonics. 
However, at $\gd=1.563\e{-4}$ and $3.125\e{-4}$ the relative contribution of higher harmonics is 
significantly larger. 
More surprisingly the frequency of the fundamental mode is the same 
for all three shear rates $\gd=7.81\e{-5}, 1.563\e{-4}$ and $3.125\e{-4}$. This is completely 
contrasted by what is observed at lower flow rates. By doubling the flow rate from 
$\gd=3.91\e{-5}$ to $\gd=7.81\e{-5}$ the position of the peaks doubles in the spectrum 
moves to the double frequency. This means that the fundamental mode at flowrate $\gd=7.81\e{-5}$ 
coincides with the the first harmonic at $\gd=3.91\e{-5}$, the first harmonic with the third and so on,
which is the expected behaviour. At higher shear rates in regime BPI-3 this lock-in of 
flow modes disappears again.

Interestingly the lock-in has only a minor effect on the flow velocities themselves.
The average velocities in Tab. \ref{tab1} reveal that only the magnitude of some of the secondary 
flow components $v_y$ and $v_z$ appears to be relatively small compared to the primary component $v_x$. 
The latter is directly linked to the imposed shear rate.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{spectrum_bp1.pdf}
\caption{Fourier analysis of the deviatoric stress $\Pi_{xy}$ of BPI: The main picture and inset show spectral
components and the lock-in of flow modes, visible in the fixed peak position for certain flow rates. 
The inset represents a blow-up of the same data. The frequency is given in radians.}
\label{bp1-spectrum}
\end{figure}

@@

@@

The rheological response of BPI at low shear rate $\gd<3.91\e{-5}$ 
is strikingly different from that of BPII and exhibits a
dynamic yield stress.
An explanation for this strange behaviour can be found by 
looking more closely at the average deviatoric stress and 
free energy density, as shown in Fig. \ref{bp1-fe-yield}.
When the quiescent and equilibrated BPI network begins to flow
the shear stress increases steeply and goes through a global maximum.
Shortly after it goes negative (the total shear stress is still
positive as the backgound viscosity has been subracted) and positive
again, forming a complete cycle. 
This cycle is repeated for a couple of times before 
the amorphous configuration emerges and any regular pattern
disappears.

\begin{figure*}[htpb]
\includegraphics[width=0.32\textwidth]{disc-xy-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-yz-400k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-700k_run1115.png}\\
\includegraphics[width=0.32\textwidth]{disc-xy-750k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-1700k_run1115.png}
\includegraphics[width=0.32\textwidth]{disc-xy-2200k_run1115.png}
\caption{Snapshots of BPI disclination network at $\dot{\gamma}=4.88\e{-6}$): 
Depicted is the transition from flow-induced, intertwined helices that 
undergo a recurring affine transformation (top left and centre at 
time step $t=4.0\e{5}$, top right at $t=7.0\e{5}$) with an intermediate 
conformation that is non-helical (bottom left at time step $t=7.5\e{5}$) 
to a distorted state (bottom centre at time step $t=1.7\e{6}$) and 
finally to an amorphous network (bottom right at $t=2.2\e{6}$). 
All pictures have the vie along the vorticity direction apart from the 
one at the top centre which shows the network in gradient direction.}
\label{bp1-low}
\end{figure*}

The negative branch in the deviatoric stress is connected
to a local maximum followed by a local minimum in the average free energy.
Hence, the region of negative stress can be 
regarded as a phase during which elastic forces excert a 
'pull' onto the network while it tries to reach a more 
favourable configuration. However, the low flow rate
causes this unstable intermediate state to be exposed to these
pulling elastic forces for a relatively long time, long enough to 
create small distortions which finally break the symmetry of the
flow-induced configuration.

\begin{figure}[htpb]
\includegraphics[width=0.495\textwidth]{stress_bp1-1_bp1-2.pdf}
\caption{Deviatoric stress $\Pi_{xy}$ over strain at the transition between regime
BPI-1 and BPI-2. The shear rates are $\gd=1.95\e{-5}$ (red solid) 
and $\gd=3.91\e{-5}$ (green dashed). Compared to lower shear rates in BPI-1 
where the stress goes negative (viz. Fig. \ref{bp1-fe-yield}) the stress is 
throughout positive, but exhibits still very large fluctuations during one cycle, 
which are absent in regime BPI-2 at higher shear rates.}
\label{bp1-1_bp1-2}
\end{figure}

@@

However, this mode of flow proves unstable as after a few cycles 
distortions appear which lead to further destabilisation.
Finally, the network loses any residual order and 
transforms into an amorphous state with virtually constant
yield stress in the region of $\Pi_{xy}\simeq 1\e{-5}$ in LBU.

@@

Adjacent to regime BPI-1 ($\gd<3.91\e{-5}$) which features yield stress 
behaviour at very low flow rates and the onset of power-law behaviour 
at slightly larger shear rates (viz. Fig. \ref{bp1-rheo}) lies another 
region ($3.91\e{-5}\le\gd< 6.25\e{-4}$) where the network flows with periodically 
recurring patterns. The transition between these two regimes is shown
in Fig. \ref{bp1-1_bp1-2} in terms of deviatoric stress over total strain.

The reason for this qualitative difference is rather obscur. A possible 
explanation is that due to the larger flow rate the network kinetically 
bypasses the relatively large stress fluctuations during one cycle which
are typical in regime BPI-1. For even lower $\gd$ the deviatoric stress
goes actually negative (viz. Fig. \ref{bp1-fe-yield}). This destabilises
the periodic network very quickly and leads to an amorphous configuration
in the steady state.

@@

The complexity of flow-induced states and the seemingly chaotic but 
periodic movement of the BPI network raises the question how robust 
these results are with respect to the system size.
Although the pitch length of 64 LBU and the equivalent of 32 LBU per
unit cell provide enough resolution to track the motion of the 
network in flow it is not clear whether 64 unit cells in total are enough to 
mimick true bulk behaviour.

@@

\begin{figure}[htpb]
\includegraphics[width=0.43\textwidth]{disc+y-600k-run911_run1163.png}\\
\includegraphics[width=0.495\textwidth]{stress_bp1_2uc_4uc.pdf}
\caption{}
\label{bp1-2uc4uc}
\end{figure}
